\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-07 chapters/gaze_independence.tex
funding

We believe it is in general not opportune to delve too deep in the underlying
ophthalmological and neurological mechanisms for each of the specific
conditions, or even the exact symptoms.
In clinical reality, every patient different symptoms, which lead to their own
unique set of preserved capabilities and visual skills.
From a solution-oriented \ac{bci} engineering point
of view, the etiology of the symptoms can be abstracted away.
Instead of categorizing patients we will refer to those that might benefit from
\ac{bci} assistive communication technology as with \ac{sspi}, in line with
the terminology used by~\textcite{FrieDOken2020}.
If they have eye motor impairment that affects their use of visual \ac{bci} or
eye tracking solutions, we use the term \ac{sspgi} (see Figure~\ref{fig:gaze_independence/venn}).

@@

\subsection{Gaze-independent visual stimulation}
This leads us to the second approach, current visual stimulation paradigms can be optimized, e.g. such that
the stimuli are always present in the field of view either overtly~\cite{Acqualagna2013, Won2018,
Lin2018} or covertly~\cite{,Pires2011,Lees2018}. Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
the GeoSpell interface~\parencite{Aloise2012}, and the \ac{rsvp}
speller~\parencite{Acqualagna2011}.
The \ac{rsvp} paradigm in particular is a prime contender for a performant gaze-independent
\ac{bci}.
\textcite{Lin2018} reported an average online \ac{itr} of 20.259 bits/min using
a paradigm with subsequently flashing sets of three characters filling the field of view of the user.
We make a distinction here between spatially organized interfaces, where
multiple targets are displayed at the same time at different spatial locations,
and temporally organized interfaces, like \ac{rsvp},
where targets or small sets of targets are shown consecutively.
The Hex-o-Spell, Geospell and GIBS Block Speller are spatially organized, while
variants of the \ac{rsvp} paradigm are temporally organized.
\textcite{Lin2018} follow the philosophy that spatially organized interface
have historically shown not to perform well in the presence of gaze impairment.
However, we argue that, spatially organized interfaces in general have a much higher
\ac{itr} in healthy controls.
For instance, a regular matrix-based speller can achive an \ac{itr} to up to
30bits/min.
With the right set of adaptations, spatial attention could potentionally also
bring added value to individuals with \ac{sspgi}.
While spatial interfaces are indeed more gaze-dependent than temporal
interfaces they also provide an extra channel that can be used to transfer
information, i.e.\ the location of the stimulus.

@@

\subsection{Gaze-independent decoding}
Thirdly, stimuli can be presented in a standard \ac{bci} paradigm, but visuospatial
attention can be decoded separately from gaze direction.
\textcite{Aloise2012} aimed to bridge the performance gap between covert and
overt VSA decoding performance.
They compared classical linear and non-linear \ac{erp} classifiers on a covert
attention oddball \ac{erp} paradigm dataset.
The results revealed no significant performance improvement in covert VSA
decoding for any of the investigated decoders.

@@

\subsubsection{The CVSA-ERP dataset}

@@

We chose a visual oddball interface that helps us study and adapt
to the effects of eye motor impairment on the \ac{erp}, when using existing state-of-the-art
decoders and on our own proposed decoders.
Using a hexagonal lay-out interface, similar to the visual Hex-o-Spell proposed
\textcite{Treder2010}, we present six flashing circular targets
(without letters or symbols) to the participant while the EEG and
electro-oculogram (EOG) are recorded, as well as eye gaze using eye tracking.
The Hex-o-Spell was chosen since it is optimiazed for gaze-independent
performance and because it has already been tested in individuals with
\ac{sspi} by~\textcite{Severens2014}.

The interface can be operated in different VSA conditions as illustrated in
Figure~\ref{fig:gaze-vsa}.

@@

  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_split.eps}
    \caption[Split \ac{vsa}]{%
      \emph{Split} \ac{vsa}.
      \Ac{vsa} is directed towards a target, while the gaze rests on another.
    }
    \label{fig:gaze/vsa/overt}
  \end{subfigure}\hfill%
   \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_free.eps}
    \caption[Free \ac{vsa}]{%
      \emph{Free} \ac{vsa}.
      Gaze and \ac{vsa} coincide on a target.
      The user is free to direct their gaze as they deem most comfortable
      All previous \ac{vsa} conditions are possible.
    }
    \label{fig:gaze/vsa/overt}
  \end{subfigure}\hfill%
  \caption[\Ac{vsa} conditions]{%
    \Ac{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm interface.
  }
  \label{fig:gaze-vsa}
\end{figure}

>>>>

We believe it is in general not opportune to delve too deep in the underlying
ophthalmological and neurological mechanisms for each of the specific
conditions, or even the exact symptoms.
In clinical reality, every patient different symptoms, which lead to their own
unique set of preserved capabilities and visual skills.
From a solution-oriented \ac{bci} engineering point
of view, the etiology of the symptoms can be abstracted away.
Instead of categorizing patients we will refer to those that might benefit from
\ac{bci} assistive communication technology as with \ac{sspi}, in line with
the terminology used by~\textcite{FriedOken2020}.
If they have eye motor impairment that affects their use of visual \ac{bci} or
eye tracking solutions, we use the term \ac{sspgi} (see Figure~\ref{fig:gaze_independence/venn}).

@@

\subsection{Gaze-independent visual stimulation}
\label{sec:gaze-independence/sota/visual}
This leads us to the second approach, current visual stimulation paradigms can be optimized, e.g. such that
the stimuli are always present in the field of view either overtly~\cite{Acqualagna2013, Won2018,
Lin2018} or covertly~\cite{,Pires2011,Lees2018}. Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
the GeoSpell interface~\parencite{Aloise2012}, and the \ac{rsvp}
speller~\parencite{Acqualagna2011}.
The \ac{rsvp} paradigm in particular is a prime contender for a performant gaze-independent
\ac{bci}.
\textcite{Lin2018} reported an average online \ac{itr} of 20.259 bits/min using
a paradigm with subsequently flashing sets of three characters filling the field of view of the user.
We make a distinction here between spatially organized interfaces, where
multiple targets are displayed at the same time at different spatial locations,
and temporally organized interfaces, like \ac{rsvp},
where targets or small sets of targets are shown consecutively.
The Hex-o-Spell, Geospell and GIBS Block Speller are spatially organized, while
variants of the \ac{rsvp} paradigm are temporally organized.
\textcite{Lin2018} follow the philosophy that spatially organized interface
have historically shown not to perform well in the presence of gaze impairment.
However, we argue that, spatially organized interfaces in general have a much higher
\ac{itr} in healthy controls.
For instance, a regular matrix-based speller can achive an \ac{itr} to up to
30bits/min.
With the right set of adaptations, spatial attention could potentionally also
bring added value to individuals with \ac{sspgi}.
While spatial interfaces are indeed more gaze-dependent than temporal
interfaces they also provide an extra channel that can be used to transfer
information, i.e.\ the location of the stimulus.

@@

\subsection{Gaze-independent decoding}
\label{sec:gaze-independence/sota/decoding}
Thirdly, stimuli can be presented in a standard \ac{bci} paradigm, but visuospatial
attention can be decoded separately from gaze direction.
\textcite{Aloise2012} aimed to bridge the performance gap between covert and
overt VSA decoding performance.
They compared classical linear and non-linear \ac{erp} classifiers on a covert
attention oddball \ac{erp} paradigm dataset.
The results revealed no significant performance improvement in covert VSA
decoding for any of the investigated decoders.

@@

\subsubsection{The CVSA-ERP dataset}
\label{sec:gaze-independence/approach/cvsa-erp}

@@

We chose a visual oddball interface that helps us study and adapt
to the effects of eye motor impairment on the \ac{erp}, when using existing state-of-the-art
decoders and on our own proposed decoders.
Using a hexagonal lay-out interface, similar to the visual Hex-o-Spell proposed
\textcite{Treder2010}, we present six flashing circular targets
(without letters or symbols) to the participant while the EEG and
electro-oculogram (EOG) are recorded, as well as eye gaze using eye tracking.
The Hex-o-Spell was chosen since it is optimized for gaze-independent
performance and because it has already been tested in individuals with
\ac{sspi} by~\textcite{Severens2014}.

The interface can be operated in different VSA conditions as illustrated in
Figure~\ref{fig:gaze/vsa}.

@@

  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_split.eps}
    \caption[Split \ac{vsa}]{%
      \emph{Split} \ac{vsa}.
      \Ac{vsa} is directed towards a target, while the gaze rests on another.
    }
    \label{fig:gaze/vsa/split}
  \end{subfigure}\hfill%
   \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_free.eps}
    \caption[Free \ac{vsa}]{%
      \emph{Free} \ac{vsa}.
      Gaze and \ac{vsa} coincide on a target.
      The user is free to direct their gaze as they deem most comfortable
      All previous \ac{vsa} conditions are possible.
    }
    \label{fig:gaze/vsa/free}
  \end{subfigure}\hfill%
  \caption[\Ac{vsa} conditions]{%
    \Ac{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm interface.
  }
  \label{fig:gaze/vsa}
\end{figure}
